---
title: "R BIEN Package: Review of Species Occurrences, Ranges, Plus finding centroids"
output: html_notebook
---

Load libraries -- tools for accessing BIEN and downloading various types of data, packages relating to maps and spatial data:

```{r}
library(BIEN)
library(maps)
library(maptools)
library(sp)
library(rgeos)

```


Quick review of something you probably already know well: accessing BIEN for species occurrence data and range data, and plotting it

```{r}
Maclura_pomifera_occ <- BIEN_occurrence_species(species = "Maclura pomifera")
Maclura_pomifera_occ
Maclura_pomifera_range <- BIEN_ranges_load_species(species = "Maclura pomifera")
map('world',fill=T , col= "grey", bg="light blue",xlim = c(-180,-20),ylim = c(-60,80))
#Now, we can add the range map:
plot(Maclura_pomifera_range,col="green",add=T)
points(cbind(Maclura_pomifera_occ$longitude,Maclura_pomifera_occ$latitude),col="red",pch=20,cex=1) 

```


I can understand that multiple range and occurrence maps will turn into a muddle. 

I do encourage you to visit ERC and if you can meet with Fatima, my understanding is that she is well versed in using R for spatial data in addition to knowing GIS. 

One option might be to use/plot centroids rather than polygons. You can calculate them in the package rgeos (already loaded in library above), with a function that can calculate the centroid of any spatial polygon, such as a range downloaded from BIEN

```{r}
centr<-gCentroid(Maclura_pomifera_range, byid = T)
centr
```

You need to modify the code used above for plotting points, because the object returned using gCentroid function uses "x" and "y" to label the values of latitude and longitude. 

There is also a reference for graphical parameters that may help you change the size and/or color of these dots if you want to do something like indicate species, or use bigger dots for the top ten species, smaller dots for the next ten, etc. 

```{r}
map('world',fill=T , col= "grey", bg="light blue",xlim = c(-180,-20),ylim = c(-60,80))
#Now, we can add the range map:
#plot(Maclura_pomifera_range,col="green",add=T)
points(cbind(centr$x,centr$y),col="red",pch=20,cex=1) 
# more graphical parameters here: https://www.statmethods.net/advgraphs/parameters.html
```
